---
title: "Lab6-wilkerson-ashley"
author: "Ashley Wilkerson"
date: "2024-11-27"
output: html_document
---


## Explaining MHV Change


```{r }
library(dplyr)
library(knitr)
library(pander)
library(stargazer)
library(scales)
library(mclust)
library( sp )          # work with shapefiles
library( sf )          # work with shapefiles - simple features format

library( tmap )        # theme maps

library( ggplot2 )     # graphing 
library( ggthemes )
library(tidycensus)
library(tigris)
library(cartogram)
set.seed(1234)



s.type <- "text"
```


```{r}
###################################
#
#     STARGAZER SETTINGS
#
###################################

# DO NOT RUN CHUNK UNLESS KNITTING:
# changes table formats to html
# before rendering RMD docs

s.type <- "html"
```


```{r}

```


## Including Plots

You can also embed plots, for example:

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits=digits)[1]
    txt <- paste(prefix, txt, sep="")
    if(missing(cex.cor)) cex <- 0.8/strwidth(txt)
    
    test <- cor.test(x,y)
    # borrowed from printCoefmat
    Signif <- symnum(test$p.value, corr = FALSE, na = FALSE,
                  cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                  symbols = c("***", "**", "*", ".", " "))
    
    text(0.5, 0.5, txt, cex = 1.5 )
    text(.7, .8, Signif, cex=cex, col=2)
}

panel.smooth <- function (x, y, col = par("col"), bg = NA, pch = par("pch"), 
    cex = 0.5, col.smooth = "red", span = 2/3, iter = 3, ...) 
{
    points(x, y, pch = 19, col = gray(0.7,0.2), bg = bg, cex = cex)
    ok <- is.finite(x) & is.finite(y)
    if (any(ok)) 
        lines(stats::lowess(x[ok], y[ok], f = span, iter = iter), 
            col = col.smooth, lwd=2, ...)
}

# custom plot
jplot <- function( x1, x2, lab1="", lab2="", draw.line=T, ... )
{

    plot( x1, x2,
          pch=19, 
          col=gray(0.6, alpha = 0.2), 
          cex=0.5,  
          bty = "n",
          xlab=lab1, 
          ylab=lab2, cex.lab=1.5,
        ... )

    if( draw.line==T ){ 
        ok <- is.finite(x1) & is.finite(x2)
        lines( lowess(x2[ok]~x1[ok]), col="red", lwd=3 ) }

}

```

## Data 
```{r}
URL1 <- "https://github.com/DS4PS/cpp-529-fall-2020/raw/main/LABS/data/rodeo/LTDB-2000.rds"
d1 <- readRDS( gzcon( url( URL1 ) ) )

URL2 <- "https://github.com/DS4PS/cpp-529-fall-2020/raw/main/LABS/data/rodeo/LTDB-2010.rds"
d2 <- readRDS( gzcon( url( URL2 ) ) )

URLmd <- "https://github.com/DS4PS/cpp-529-fall-2020/raw/main/LABS/data/rodeo/LTDB-META-DATA.rds"
md <- readRDS( gzcon( url( URLmd ) ) )

d1 <- select( d1, - year )
d2 <- select( d2, - year )

d <- merge( d1, d2, by="tractid" )
d <- merge( d, md, by="tractid" )
```

## Filter Rural Districts

```{r}
table( d$urban )
```

```{r}
d <- filter ( d, urban == 'urban')
```

## Create New Variables

```{r}
d <- select( d, tractid, 
             mhmval00, mhmval12, 
             hinc00, 
             hu00, vac00, own00, rent00, h30old00,
             empclf00, clf00, unemp00, prof00,  
             dpov00, npov00,
             ag25up00, hs00, col00, 
             pop00.x, nhwht00, nhblk00, hisp00, asian00,
             cbsa, cbsaname )

 
d <- 
  d %>%
  mutate( # percent white in 2000
          p.white = 100 * nhwht00 / pop00.x,
          # percent black in 2000
          p.black = 100 * nhblk00 / pop00.x,
          # percent hispanic in 2000
          p.hisp = 100 * hisp00 / pop00.x, 
          # percent asian in 2000
          p.asian = 100 * asian00 / pop00.x,
          # percent high school grads by age 25 in 2000 
          p.hs = 100 * (hs00+col00) / ag25up00,
          # percent pop with college degree in 2000
          p.col = 100 * col00 / ag25up00,
          # percent employed in professional fields in 2000
          p.prof = 100 * prof00 / empclf00,
          # percent unemployment  in 2000
          p.unemp = 100 * unemp00 / clf00,
          # percent of housing lots in tract that are vacant in 2000
          p.vacant = 100 * vac00 / hu00,
          # dollar change in median home value 2000 to 2010 
          pov.rate = 100 * npov00 / dpov00 )


# adjust 2000 home values for inflation 
mhv.00 <- d$mhmval00 * 1.28855  
mhv.10 <- d$mhmval12

# change in MHV in dollars
mhv.change <- mhv.10 - mhv.00


# drop low 2000 median home values
# to avoid unrealistic growth rates.
#
# tracts with homes that cost less than
# $1,000 are outliers
mhv.00[ mhv.00 < 1000 ] <- NA

# change in MHV in percent
mhv.growth <- 100 * ( mhv.change / mhv.00 )

d$mhv.00 <- mhv.00
d$mhv.10 <- mhv.10
d$mhv.change <- mhv.change
d$mhv.growth <- mhv.growth
```

## MEDIAN HOME VALUE
```{r}
hist( mhv.00, breaks=200, xlim=c(0,500000), 
      col="gray20", border="white",
      axes=F, 
      xlab="MHV (median = $138k)",
      ylab="",
      main="Median Home Value in 2000 (2010 US dollars)" )

axis( side=1, at=seq(0,500000,100000), 
      labels=c("$0","$100k","$200k","$300k","$400k","$500k") )

abline( v=median( mhv.00, na.rm=T ), col="orange", lwd=3 )
```

## DESCRIPTIVES

```{r}
df <- data.frame( MedianHomeValue2000=mhv.00, 
                  MedianHomeValue2010=mhv.10, 
                  MHV.Change.00.to.10=mhv.change,
                  MHV.Growth.00.to.12=mhv.growth )
s.type <- "text"
stargazer( df, 
           type=s.type, 
           digits=0, 
           summary.stat = c("min", "p25","median","mean","p75","max") )
```

## CHANGE IN MHV 2000-2010
```{r}
hist( mhv.change/1000, breaks=500, 
      xlim=c(-100,500), yaxt="n", xaxt="n",
      xlab="Thousand of US Dollars (adjusted to 2010)", cex.lab=1.5,
      ylab="", main="Change in Median Home Value 2000 to 2010",
      col="gray20", border="white" )

axis( side=1, at=seq( from=-100, to=500, by=100 ), 
      labels=paste0( "$", seq( from=-100, to=500, by=100 ), "k" ) )
        
mean.x <- mean( mhv.change/1000, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=200, y=1500, 
      labels=paste0( "Mean = ", dollar( round(1000*mean.x,0)) ), 
      col="darkorange", cex=1.8, pos=3 )

median.x <- median( mhv.change/1000, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=200, y=2000, 
      labels=paste0( "Median = ", dollar( round(1000*median.x,0)) ), 
      col="dodgerblue", cex=1.8, pos=3 )
```

## PERCENT CHANGE IN MHV 2000 TO 2010 

```{r}
hg <-
hist( mhv.growth, breaks=5000, 
      xlim=c(-100,200), yaxt="n", xaxt="n",
      xlab="", cex.main=1.5,
      ylab="", main="Growth in Home Value by Census Tract 2000 to 2010",
      col="gray40", border="white" )

axis( side=1, at=seq( from=-100, to=200, by=50 ), 
      labels=paste0( seq( from=-100, to=200, by=50 ), "%" ) )

ymax <- max( hg$count )
        
mean.x <- mean( mhv.growth, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=100, y=(0.5*ymax), 
      labels=paste0( "Mean = ", round(mean.x,0), "%"), 
      col="darkorange", cex=1.8, pos=4 )

median.x <- median( mhv.growth, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=100, y=(0.6*ymax), 
      labels=paste0( "Median = ", round(median.x,0), "%"), 
      col="dodgerblue", cex=1.8, pos=4 )
```

## METRO LEVEL STATISTICS

```{r}
# average growth in median home value for the city
d <- 
  d %>%
  group_by( cbsaname ) %>%
  mutate( metro.mhv.change = median( mhv.change, na.rm=T ),
             metro.mhv.growth = 100 * median( mhv.growth, na.rm=T ) ) %>%
  ungroup()
```

##  VARIABLE SELECTION

```{r}
colnames(d)
```

> Percentage of black residents, Hypotheses: in areas with a higher percentage of black residents in literature median home values seen to go down..

```{r}
par( mfrow=c(1,2) )

hist( d$p.black, breaks=50, col="gray20", border="white",
      yaxt="n", xlab="", ylab="", main="Percentage Black Residents")
hist( log(d$p.black+1), breaks=50, col="gray20", border="white",
      yaxt="n", xlab="", ylab="", main="Percentage Black Residents (logged)")
```

>Bachelor Degree in 2000’s ###Hypothesis: Higher levels of education in 2000 will predict a larger increase in home value between 2000 and 2010.

```{r}
hist( d$col00, breaks=50, col="gray20", border="white",
      yaxt="n", xlab="", ylab="", main="College Degree")
```

```{r}
hist( log(d$col00+1), breaks=50, col="gray20", border="white",
      yaxt="n", xlab="", ylab="", main="College Degree (logged)")
```

>Percent Unemployed ###Hypothesis: Higher levels of unemployment will predict a larger gap in median home prices between 2000 and 2010.

```{r}
hist( d$p.unemp, breaks=50, col="gray20", border="white",
      yaxt="n", xlab="", ylab="", main="Percent Unemployed")
```

```{r}
hist( log(d$p.unemp+1), breaks=50, col="gray20", border="white",
      yaxt="n", xlab="", ylab="", main="Percent Unemployed (logged)")
```

```{r}
log.p.unemp <- log10( d$p.unemp + 1 )
log.p.black <- log10( d$p.black + 1 )
log.col00 <- log10( d$col00 + 1)
these <- sample( 1:length(log.p.unemp), 5000 )

par( mfrow=c(1,2) )
jplot( d$p.unemp[these], d$p.black[these], 
       lab1="Percentage Unemployed", lab2="Percentage Black Residents",
       main="Raw Measures" )
jplot( log.p.unemp[these], log.p.black[these], 
       lab1="Percentage Unemployed", lab2="Percentage Black Residents",
       main="Log Transformed" )
```

```{r}
cor( d$p.unemp, d$p.black, use="pairwise.complete" )
```

```{r}
cor( log.p.unemp, log.p.black, use="pairwise.complete" )
```

```{r}
# create subset to visualize in correlation matrix 
d2 <- select( d, mhv.growth, p.black,  p.prof,  pov.rate, p.unemp )

# reduce data density for visualization
set.seed( 1234 )
d3 <- sample_n( d2, 10000 ) %>% na.omit()

# correlation plots

pairs( d3, upper.panel=panel.cor, lower.panel=panel.smooth)
```

```{r}
jplot( d3$p.black, d3$mhv.growth, ylim=c(-50,100),
       lab1="Vacancy Rates", lab2="MHV Growth" )
```

```{r}
reg.data <- d

reg.data$mhv.growth[ reg.data$mhv.growth > 200 ] <- NA
reg.data$p.unemp <- log10( reg.data$p.unemp + 1 )
reg.data$p.black <- log10( reg.data$p.black + 1 )
reg.data$col00 <- log10( reg.data$col00 + 1 )

m1 <- lm( mhv.growth ~ p.black, data=reg.data )
m2 <- lm( mhv.growth ~  p.unemp, data=reg.data )
m3 <- lm( mhv.growth ~  col00 + p.unemp, data=reg.data )


stargazer( m1, m2, m3, 
           type=s.type, digits=2,
           omit.stat = c("rsq","f") )
```

```{r}
# Calculate changes
d$delta_p_black <- d$p.black - (d$nhblk00 / d$pop00.x)
d$delta_p_unemp <- d$p.unemp - (d$unemp00 / d$clf00)
d$delta_p_col <- d$p.col - (d$col00 / d$ag25up00)


# Regression models with changes in covariates
m1_change <- lm(mhv.growth ~ delta_p_black, data = d)
m2_change <- lm(mhv.growth ~ delta_p_unemp, data = d)
m3_change <- lm(mhv.growth ~ delta_p_col + delta_p_unemp, data = d)

# Compare results using stargazer
library(stargazer)

stargazer(m1_change, m2_change, m3_change,
          type = "text", digits = 2,
          omit.stat = c("rsq", "f"),
          title = "Regression Results Using Changes in Covariates")


```


```{r}
census_api_key("0765edf09e6abf32a5af72fafd38dd94374a38c4")

crosswalk <- read.csv( "https://raw.githubusercontent.com/DS4PS/cpp-529-master/master/data/cbsatocountycrosswalk.csv",  stringsAsFactors=F, colClasses="character" )
```


```{r}
grep( "^NEW YORK", crosswalk$msaname, value=TRUE )
```

```{r}
these.msp <- crosswalk$msaname == "NEW YORK"
these.fips <- crosswalk$fipscounty[ these.msp ]
these.fips <- na.omit( these.fips )

head( these.fips ) %>% pander()
```

```{r}
state.fips <- substr( these.fips, 1, 2 )
county.fips <- substr( these.fips, 3, 5 )

cbind( these.fips, state.fips, county.fips ) %>% pander()
```

```{r}
valid_fips <- county.fips[state.fips == "36"]
valid_fips <- valid_fips[valid_fips != "990"]

msp.pop1 <- get_acs(
  geography = "tract",
  variables = "B01003_001",  # Total population
  state = "36",              # New York State FIPS code
  county = county.fips[state.fips == "36"],  # Filter counties within New York State
  geometry = TRUE             
) %>% 
  select(GEOID, estimate) %>%  
  rename(POP = estimate)       # Rename estimate column to POP
```


```{r}
URL <- "https://github.com/DS4PS/cpp-529-master/raw/master/data/ltdb_std_2010_sample.rds"
census.dat <- readRDS(gzcon(url(URL)))

# Merge spatial data (msp.pop) with census data
msp <- msp.pop1 %>%
  left_join(census.dat, by = c("GEOID" = "tractid"))

# Remove empty polygons
msp <- msp[!st_is_empty(msp), ]
```

```{r}
plot( msp.pop1 )
```

```{r}
dd.URL <- "https://raw.githubusercontent.com/DS4PS/cpp-529-master/master/data/data-dictionary.csv"
data.dictionary <- read.csv( dd.URL, stringsAsFactors=F ) 
data.dictionary %>% pander()
```


```{r}
# project map and remove empty tracts
msp <- st_transform( msp, CRS("+init=epsg:3395"))
```

```{r}
msp <- msp[ msp$POP != 0 & (! is.na( msp$POP )) , ]

# convert census tract polygons to dorling cartogram
# no idea why k=0.03 works, but it does - default is k=5
msp$pop.w <- msp$POP / 9000 # max(msp.sp$POP)   # standardizes it to max of 1.5
msp_dorling <- cartogram_dorling( x=msp, weight="pop.w", k=0.05 )
plot( msp_dorling )
```


```{r}
# Ensure interactive or static mode as needed
tmap_mode("plot")  # or "view" for interactive maps
```


```{r}
# Plot the Dorling cartogram with color based on "hinc12"
tm_shape(msp_dorling) + 
  tm_polygons(col = "hinc12", n = 7, style = "quantile", palette = "Spectral") +
  tm_layout(title = "Dorling Cartogram", title.position = c("right", "top"))
```

```{r}
bb <- st_bbox( c( xmin =  -10451823, xmax = -10324525, 
                  ymax = 5639769, ymin = 5491665 ), 
               crs = st_crs("+init=epsg:3395"))

tm_shape(msp_dorling, bbox = bb) + 
  tm_polygons(col = "hinc12", n = 10, style = "quantile", palette = "Spectral") +
  tm_layout(title = "Dorling Cartogram", title.position = c("right", "top"))
```

```{r}
keep.these <- c("pnhwht12", "pnhblk12", "phisp12", "pntv12", "pfb12", "polang12", 
"phs12", "pcol12", "punemp12", "pflabf12", "pprof12", "pmanuf12", 
"pvet12", "psemp12", "hinc12", "incpc12", "ppov12", "pown12", 
"pvac12", "pmulti12", "mrent12", "mhmval12", "p30old12", "p10yrs12", 
"p18und12", "p60up12", "p75up12", "pmar12", "pwds12", "pfhh12")

d1 <- st_drop_geometry(msp_dorling)
d2 <- select( d1, keep.these )
```

```{r}
d3 <- apply( d2, 2, scale )
head( d3[,1:6] ) %>% pander()
```

```{r}
keep.these <- c("pnhwht12", "pnhblk12", "phisp12", "pntv12", "pfb12", "polang12", 
"phs12", "pcol12", "punemp12", "pflabf12", "pprof12", "pmanuf12", 
"pvet12", "psemp12", "hinc12", "incpc12", "ppov12", "pown12", 
"pvac12", "pmulti12", "mrent12", "mhmval12", "p30old12", "p10yrs12", 
"p18und12", "p60up12", "p75up12", "pmar12", "pwds12", "pfhh12")

d2 <- select( d1, keep.these )
d3 <- apply( d2, 2, scale )

head( d3[,1:6] ) %>% pander()
```

```{r}
library(dplyr)
library(sf)
library(mclust)
library(pander)

# Define columns to keep
keep.these <- c("pnhwht12", "pnhblk12", "phisp12", "pntv12", "pfb12", "polang12", 
                "phs12", "pcol12", "punemp12", "pflabf12", "pprof12", "pmanuf12", 
                "pvet12", "psemp12", "hinc12", "incpc12", "ppov12", "pown12", 
                "pvac12", "pmulti12", "mrent12", "mhmval12", "p30old12", "p10yrs12", 
                "p18und12", "p60up12", "p75up12", "pmar12", "pwds12", "pfhh12")


d1 <- st_drop_geometry(msp_dorling)
d2 <- select(d1, all_of(keep.these))

rows_to_keep <- complete.cases(d2)

d3 <- apply(d2[rows_to_keep, ], 2, scale)


msp_dorling <- msp_dorling[rows_to_keep, ]

set.seed(1234)
fit <- Mclust(d3)
msp_dorling$cluster <- as.factor(fit$classification)

summary(fit)
```























